Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-14 article.tex
clean up article section on error reporting parser

\section{Example 2: error-reporting}
While our first attempt at a parser was useful, when it encountered malformed input, it wasn't
helpful -- all it could do was throw up its hands in despair and give up.  However, we'd like
an error message that tells us what the problem was and where it was encountered whenever bad
input is encountered.  Otherwise, we'd waste time finding the problem by eye -- not a lot of 
fun.  Here are the examples of bad input that we'd like to get specific error messages for:
\begin{itemize}
\item application: missing operator:  \begin{verbatim}()\end{verbatim}
\item application: missing close parenthesis:  \begin{verbatim}(a b\end{verbatim}
\item define: missing symbol:  \begin{verbatim}{define (a b c)}\end{verbatim}
\item define: missing form:  \begin{verbatim}{define a}\end{verbatim}
\item define: missing close curly:  \begin{verbatim}{define a b\end{verbatim}
\item lambda: missing parameter list:  \begin{verbatim}{lambda (a b c)}\end{verbatim}
\item lambda: duplicate parameter names:  \begin{verbatim}{lambda {a b a} (c d)}\end{verbatim}
\item lambda: missing parameter list close curly:  \begin{verbatim}{lambda {a b (c d)}\end{verbatim}
\item lambda: missing body form:  \begin{verbatim}{lambda {a b}}\end{verbatim}
\item lambda: missing close curly:  \begin{verbatim}{lambda {a b} (c d)\end{verbatim}
\item special form: unable to parse:  \begin{verbatim}{defin x y}\end{verbatim}
\item woof: unparsed input:  \begin{verbatim}a,b\end{verbatim}
\end{itemize}
To do this, we're going to bring in a second monad transformer type class -- MonadError -- and
another monad transformer datatype -- MaybeT.  Our parser stack and runParser function are now
going to be:
\begin{verbatim}
type Parse e t a = StateT [t] (MaybeT (Either' e)) a

runParser :: Parse e t a -> [t] -> Either' e (Maybe (a, [t]))
runParser p xs = runMaybeT (runStateT p xs)
\end{verbatim}
No, that ' isn't accidental -- we again introduce a replacement data type to avoid a type class
instance that doesn't fulfill our needs.  In this case, Either' is intended as a replacement for
Either, with the only difference being its instance of MonadError.  All other instances for 
Either' are analagous to those of Either:
\begin{verbatim}
data Either' a b
    = Left' a
    | Right' b
  deriving (Show, Eq)

instance MonadError e (Either' e) where
  throwError                =  Left'
  catchError  (Right' x) _  =  Right' x
  catchError  (Left' e)  f  =  f e
\end{verbatim}
We'll also need an instance of Switch for MaybeT:
\begin{verbatim}
instance Functor m => Switch (MaybeT m) where
  switch (MaybeT m) = MaybeT (fmap switch m)
\end{verbatim}

@@

\subsection{Token parsers}
Unfortunately, by adding the position to the input, we've created a new problem for ourselves: our 
original token parsers assumed they would be operating on character lists, so we'll have to rewrite
them to get rid of that assumption.  The cause of the problem is the \verb+literal+ combinator, which
we can easily replace with this one that assumes token lists:
\begin{verbatim}
character :: (MonadState [Token] m, Plus m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}
And now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 (not1 $ character '\n')

@@

ocurly = munch $ character '{'
ccurly = munch $ character '}'
oparen = munch $ character '('
cparen = munch $ character ')'
symbol = munch $ many1 char
  where char = fmap chr $ satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']) . chr)
\end{verbatim}
The only differences are that literal was replaced everywhere by character, and a couple uses
of the chr accessor.  The munch combinator didn't have to change at all (although its inferred
type did in fact change).

\subsection{Error messages}
Now we get to the meat of this section -- modifying the parsers to report accurate and useful errors.
Let's start out with a list, in code, of all the errors that we're going to deal with:
\begin{verbatim}
eaOp   =  "application: missing operator"
eaCls  =  "application: missing close parenthesis"
edSym  =  "define: missing symbol"
edForm =  "define: missing form"
edCls  =  "define: missing close curly"
elPL   =  "lambda: missing parameter list"
elPrms =  "lambda: duplicate parameter names"
elPCls =  "lambda: missing parameter list close curly"
elBody =  "lambda: missing body form"
elCls  =  "lambda: missing close curly"
esName =  "special form: unable to parse"
ewUnp  =  "woof: unparsed input"
\end{verbatim}
The MonadError type class provides two useful combinators, throwError for generating errors, and 
catchError for dealing with them.  We can ignore catchError because we don't want to try to recover
from errors, we just want them reported.  Now the question is, when should an error be created
and what data should it include?  \verb+throwError+ is too basic for our needs, so we'll introduce
the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Plus m) => e -> m a -> m a
commit err p = p <+> throwError err
\end{verbatim}
This combinator takes two arguments: 1. an error value, and 2. a monadic computation.  It first tries
to run the computation, and if the computation fails, generates an error whose contents are the first
argument.  The definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+.
So what data do we want to include in our errors?  A String describing the error and a token reporting
the position should describe the error decently:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}
Now lets start with \verb+Application+ as our first modified, error-reporting parser:
\begin{verbatim}
application =
    oparen                       >>= \open ->
    commit (eaOp, open) form     >>= \op ->
    many0 form                   >>= \args ->
    commit (eaCls, open) cparen  >>
    return (AApp op args)
\end{verbatim}
This parser works by running the open-parenthesis parser, capturing its value in \verb+open+, and continuing
with the parsing; if either \verb+form+ or close-parenthesis fails, an appropriate error can be generated
which includes a string and the location of the open parenthesis.  However, if the open parenthesis is not
successfully parsed, it's a failure but not an error.  Here are some examples to demonstrate this in action:
\begin{verbatim}
*Main> runParser application (countLineCol "oops")
Right' Nothing

*Main> runParser application (countLineCol "(a b c)")
Right' (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))

*Main> runParser application (countLineCol "\n()")
Left' ("application: missing operator",('(',2,1))

*Main> runParser application (countLineCol "(a b")
Left' ("application: missing close parenthesis",('(',1,1))
\end{verbatim}
Although the previous version of the \verb+application+ parser could be implemented with just applicative combinators, 
this version certainly can't:  \verb+open+ is used to build the following parsers, making this decidedly monadic.

The \verb+define+ and \verb+lambda+ parsers are upgraded similarly:
\begin{verbatim}
define =
    ocurly                       >>= \open ->
    check (== "define") symbol    *>
    pure ADefine                 <*>
    commit (edSym, open) symbol  <*>
    commit (edForm, open) form   <*
    commit (edCls, open) ccurly

lambda =
    ocurly                               >>= \open ->
    check (== "lambda") symbol           >>
    commit (elPL, open) ocurly           >>= \p_open ->
    many0 symbol                         >>= \params ->
    (if distinct params 
        then return ()
        else throwError (elPrms, open))  >>
    commit (elPCls, open) ccurly         >>
    commit (elBody, open) (many1 form)   >>= \bodies ->
    commit (elCls, open) ccurly          >>
    return (ALambda params bodies)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
Note that \verb+commit+ isn't used until we can be sure that we're in the right rule:  if, in the \verb+define+
parser, we committed to parsing the "define" keyword, we wouldn't be able to backtrack and try the \verb+lambda+
parser if that failed.  Thus, if treating errors as unrecoverable, it's important not to place a \verb+commit+
where backtracking might be necessary to parse a valid alternative.  However, this isn't the case with how
we've used it, since once we see an open curly brace and a define, we're sure that we're in the \verb+define+
rule, and allowing backtracking would only destroy our ability to produce clean and accurate error reports.

@@

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
runParser woof (countLineCol "{define fgh {lambda {f x y} {define a (b c)} (f y a (a x))}}")
Right' (Just ([ADefine "fgh" 
                       (ALambda ["f","x","y"] 
                                [ADefine "a" (AApp (ASymbol "b") [ASymbol "c"]),
                                 AApp (ASymbol "f") [ASymbol "y",
                                                     ASymbol "a",
                                                     AApp (ASymbol "a") [ASymbol "x"]]])],
              []))
\end{verbatim}
Here are examples showing errors detected by the parser:
\begin{verbatim}
*Main> map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
                                     "{define a}", "{define a b", 
                                     "{lambda (a b c)}", "{lambda {a b a} (c d)}", 
                                     "{lambda {a b (c d)}", "{lambda {a b}}", 
                                     "{lambda {a b} (c d)", "{defin x y}", "a,b"]
Left ("application: missing operator",('(',1,1))
Left ("application: missing close parenthesis",('(',1,1))
Left ("define: missing symbol",('{',1,1))
Left ("define: missing form",('{',1,1))
Left ("define: missing close curly",('{',1,1))
Left ("lambda: missing parameter list",('{',1,1))
Left ("lambda: duplicate parameter names",('{',1,1))
Left ("lambda: missing parameter list close curly",('{',1,1))
Left ("lambda: missing body form",('{',1,1))
Left ("lambda: missing close curly",('{',1,1))
Left ("special form: unable to parse",('{',1,1))
Left ("woof: unparsed input",(',',1,2))
\end{verbatim}

>>>>

\section{Example 2: error-reporting}
While our first parser worked great when the input was valid, it wasn't helpful
at all when the input was malformed.  We'd like it to produce an error message
that tells us what the problem was and where it was encountered whenever bad
input is encountered.  Otherwise, we'd waste time finding the problem by eye 
-- not a lot of fun.

To solve this problem, let's start with an informal spec outlining the error
messages that we expect to get, and the malformed input that causes each of them: 
\begin{itemize}
\item application: missing operator:  \begin{verbatim}()\end{verbatim}
\item application: missing close parenthesis:  \begin{verbatim}(a b\end{verbatim}
\item define: missing symbol:  \begin{verbatim}{define (a b c)}\end{verbatim}
\item define: missing form:  \begin{verbatim}{define a}\end{verbatim}
\item define: missing close curly:  \begin{verbatim}{define a b\end{verbatim}
\item lambda: missing parameter list:  \begin{verbatim}{lambda (a b c)}\end{verbatim}
\item lambda: duplicate parameter names:  \begin{verbatim}{lambda {a b a} (c d)}\end{verbatim}
\item lambda: missing parameter list close curly:  \begin{verbatim}{lambda {a b (c d)}\end{verbatim}
\item lambda: missing body form:  \begin{verbatim}{lambda {a b}}\end{verbatim}
\item lambda: missing close curly:  \begin{verbatim}{lambda {a b} (c d)\end{verbatim}
\item special form: unable to parse:  \begin{verbatim}{defin x y}\end{verbatim}
\item woof: unparsed input:  \begin{verbatim}a,b\end{verbatim}
\end{itemize}
To do this, we're going to bring in a second monad transformer type class -- MonadError -- and
another monad transformer datatype -- MaybeT.  Our parser stack and runParser function are now:
\begin{verbatim}
type Parse e t a = StateT [t] (MaybeT (Either e)) a

runParser :: Parse e t a -> [t] -> Either e (Maybe (a, [t]))
runParser p xs = runMaybeT (runStateT p xs)
\end{verbatim}
Now we have three different possible results:  success, failure, and an error:
\begin{verbatim}
*Main> runParser item "abc"
Right (Just ('a',"bc"))

*Main> runParser item ""
Right Nothing

*Main> runParser (throwError "oops!") ""
Left "oops!"
\end{verbatim}
Although the choice combinator can recover from failures, it can't recover from
errors.  Compare:
\begin{verbatim}
*Main> runParser (satisfy (== 'a') <|> satisfy (== 'b')) "babc"
Right (Just ('b',"abc"))

*Main> runParser (throwError "no backtracking!" <|> satisfy (== 'b')) "babc"
Left "no backtracking!"
\end{verbatim}
We'll take advantage of this backtracking restriction to produce accurate error
messages.

\subsection{Type Class Preliminaries}
Unfortunately, the standard MonadError instance for the Either datatype does
not fit our needs -- it requires an Error constraint that we do not want.  To
solve this problem, we'll implement our own MonadError type class, based on the
standard one, and instances of it.  We'll need an instance for each member of
our stack -- Either, StateT, and MaybeT: 
\begin{verbatim}
class Monad m => MonadError e m | m -> e where
  throwError :: e -> m a
  catchError :: m a -> (e -> m a) -> m a

instance MonadError e (Either e) where
  throwError               =  Left
  catchError  (Right x) _  =  Right x
  catchError  (Left e)  f  =  f e

instance MonadError e m => MonadError e (StateT s m) where
  throwError      =  lift . throwError
  catchError m f  =  StateT (\s -> catchError (runStateT m s) 
                                              (\e -> runStateT (f e) s))

instance MonadError e m => MonadError e (MaybeT m) where
  throwError      =  lift . throwError
  catchError m f  =  MaybeT $ catchError (runMaybeT m) (runMaybeT . f)
\end{verbatim}
Note how the StateT and MaybeT instances don't really do anything with the
errors; they just pass the error on through to the next level.  That's why they
both require that the next lower level supports MonadError.  Also, this code 
requires some additional extensions:
\begin{itemize}
\item FlexibleInstances
\item FunctionalDependencies
\item UndecidableInstances 
\end{itemize}

We'll also need an instance of Switch for MaybeT:
\begin{verbatim}
instance Functor m => Switch (MaybeT m) where
  switch (MaybeT m) = MaybeT (fmap switch m)
\end{verbatim}

@@

\subsection{Token parsers}
By adding the position to the input, we've created a new problem 
for ourselves: our original token parsers operated on 
character lists, but we now need them to work on token lists.  
The cause of the problem is the \verb+literal+ combinator, so we
replace it with one for token lists:
\begin{verbatim}
character :: (MonadState [Token] m, Plus m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}
And now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 (not1 $ character '\n')

@@

ocurly = munch $ character '{'
ccurly = munch $ character '}'
oparen = munch $ character '('
cparen = munch $ character ')'
symbol = munch $ many1 char
  where char = fmap chr $ satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']) . chr)
\end{verbatim}
The only differences were the replacement of literal with character and a couple
uses of the chr accessor.  The munch combinator didn't have to change at all 
(although its inferred type did in fact change).

\subsection{Error messages}
Now we modify the remaining parsers to report accurate and useful errors.
Here's the error messages that we'll use to report each error:
\begin{verbatim}
eAppOper    =  "application: missing operator"
eAppClose   =  "application: missing close parenthesis"
eDefSym     =  "define: missing symbol"
eDefForm    =  "define: missing form"
eDefClose   =  "define: missing close curly"
eLamParam   =  "lambda: missing parameter list"
eLamDupe    =  "lambda: duplicate parameter names"
eLamPClose  =  "lambda: missing parameter list close curly"
eLamBody    =  "lambda: missing body form"
eLamClose   =  "lambda: missing close curly"
eSpecial    =  "special form: unable to parse"
eWoof       =  "woof: unparsed input"
\end{verbatim}

The MonadError type class provides two useful combinators, throwError for 
generating errors, and catchError for dealing with them.  We can ignore 
catchError because we don't need to recover from errors, just report them. 
When should an error be generated and what 
data should it include?  \verb+throwError+ is too basic for our needs, so we'll 
introduce the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Plus m) => e -> m a -> m a
commit err p = p <+> throwError err
\end{verbatim}
This combinator takes two arguments: 1. an error value, and 2. a monadic 
computation.  It first tries to run the computation, and if the computation 
fails, generates an error whose contents are the first argument.  The 
definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+. So 
what data do we want to include in our errors?  A String describing the error 
and a token reporting the position should describe the error decently:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}
Now lets start with \verb+Application+ as our first modified, error-reporting parser:
\begin{verbatim}
application =
    openparen                             >>= \open ->
    commit (eAppOper, open) form          >>= \op ->
    many0 form                            >>= \args ->
    commit (eAppClose, open) closeparen   >>
    return (AApp op args)
\end{verbatim}
This parser works by running the open-parenthesis parser, capturing its value 
in \verb+open+, and continuing with the parsing; if either \verb+form+ or 
close-parenthesis fails, an appropriate error can be generated which includes 
a string and the location of the open parenthesis.  However, if the open 
parenthesis is not successfully parsed, it's a failure but not an error.  
Here are some examples to demonstrate this in action:
\begin{verbatim}
*Main> runParser application (countLineCol "oops")
Right Nothing

*Main> runParser application (countLineCol "(a b c)")
Right (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))

*Main> runParser application (countLineCol "\n()")
Left ("application: missing operator",('(',2,1))

*Main> runParser application (countLineCol "(a b")
Left ("application: missing close parenthesis",('(',1,1))
\end{verbatim}
Although the previous version of the \verb+application+ parser was implemented 
with just applicative combinators, this version can't be:  \verb+open+ 
is used to build the following parsers, making this decidedly monadic.

The \verb+define+ and \verb+lambda+ parsers are upgraded similarly:
\begin{verbatim}
define =
    opencurly                            >>= \open ->
    check (== "define") symbol            *>
    pure ADefine                         <*>
    commit (eDefSym, open) symbol        <*>
    commit (eDefForm, open) form         <*
    commit (eDefClose, open) closecurly

lambda =
    opencurly                                >>= \open ->
    check (== "lambda") symbol               >>
    commit (eLamParam, open) opencurly       >>= \p_open ->
    many0 symbol                             >>= \params ->
    (if distinct params 
        then return ()
        else throwError (eLamDupe, p_open))  >>
    commit (eLamPClose, p_open) closecurly   >>
    commit (eLamBody, open) (many1 form)     >>= \bodies ->
    commit (eLamClose, open) closecurly      >>
    return (ALambda params bodies)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
Note that \verb+commit+ isn't used until we can be sure that we're in the right 
rule:  if, in the \verb+define+ parser, we committed to parsing the "define" 
keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
that failed.  Thus, if treating errors as unrecoverable, it's important not to 
place a \verb+commit+ where backtracking might be necessary to parse a valid 
alternative.  However, this isn't the case with how we've used it, since once 
we see an open curly brace and a define, we're sure that we're in the \verb+define+
rule, and allowing backtracking would only destroy our ability to produce clean 
and accurate error reports.

@@

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
*Main> runParser woof (countLineCol "{define fgh {lambda {f x} (f x)}}")
Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
                                     [AApp (ASymbol "f") [ASymbol "x"]])],
       []))
\end{verbatim}
Here's the proof showing that the parser correctly report all of the errors 
covered by our spec:
\begin{verbatim}
*Main> let p = runParser woof . countLineCol

*Main> p "()"
Left ("application: missing operator",('(',1,1))

*Main> p "(a b"
Left ("application: missing close parenthesis",('(',1,1))

*Main> p "{define (a b c)}"
Left ("define: missing symbol",('{',1,1))

*Main> p "{define a}"
Left ("define: missing form",('{',1,1))

*Main> p "{define a b"
Left ("define: missing close curly",('{',1,1))

*Main> p "{lambda (a b c)}"
Left ("lambda: missing parameter list",('{',1,1))

*Main> p "{lambda {a b a} (a b c)}"
Left ("lambda: duplicate parameter names",('{',1,9))

*Main> p "{lambda {a b (c d)}"
Left ("lambda: missing parameter list close curly",('{',1,9))

*Main> p "{lambda {a b}}"
Left ("lambda: missing body form",('{',1,1))

*Main> p "{lambda {a b} (c d)"
Left ("lambda: missing close curly",('{',1,1))

*Main> p "{defin x y}"
Left ("special form: unable to parse",('{',1,1))

*Main> p "a,b"
Left ("woof: unparsed input",(',',1,2))
\end{verbatim}
